---
title: "Census Pull"
author: "Ethan Tenison"
date: "`r format(Sys.Date(), '%B %d, %Y') `"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(tidyverse)
library(tidycensus)
```


# ACS Variables 

* Wealth
  + QRICH = Percent Households Earning over $200,000 annually
  + MDHSEVAL = Median Housing Value
  + PERCAP = Per Capita Income
* Language & Education 
  + QESL = Percent Speaking English as a Second Language with Limited Proficiency
  + QSPANISH =  Percent Hispanic 
  + QED12LES = Percent Less than high school education for population over 25 years and older
* Elderly
  + QSSBEN = Percent Households Receiving Social Security Benefits
  + QAGEDEP = Percent Population under 5 years or 65 and over
  + MEDAGE = Median age 
* Housing Status 
  + PPUNIT = People per Unit (Average household size)
  + QFAM = Percent Children Living in Married Couple Families
* Social Status
  + QCVLUN = Percent Unemployment for Civilian in Labor Force 16 Years and Over
  + QBLACK = Percent Black or African American Alone
  + QNOAUTO = Percent Housing Units with No Car
  + QPOVTY = Percent Poverty
* Gender
  + QFEMALE = Percent Female 
  + QFEMLBR = Percent Female Participation in Labor Force


```{r var}

#acs variables 
vars <- c("B02001_001", "B02001_002", "B02001_003", "B02001_005",
          "B03002_012", "B06012_002", "B06012_003")

counties <- c("Travis", "Bastrop","Blanco", "Burnet", "Caldwell",
              "Fayette", "Hays", "Lee", "Llano", "Williamson")

#enter your key census_api_key("Your key here")
acs <-get_acs(state="TX", county = counties, geography="tract",year = 2019,
              variables=vars, geometry= F) 

```
