---
title: "Census Pull"
author: "Ethan Tenison"
date: "`r format(Sys.Date(), '%B %d, %Y') `"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(tidyverse)
library(tidycensus)
```


# ACS Variables 

* Wealth
  + QRICH = Percent Households Earning over $200,000 annually
  + MDHSEVAL = Median Housing Value
  + PERCAP = Per Capita Income
* Language & Education 
  + QESL = Percent Speaking English as a Second Language with Limited Proficiency
  + QSPANISH =  Percent Hispanic 
  + QED12LES = Percent Less than high school education for population over 25 years and older
* Elderly
  + QSSBEN = Percent Households Receiving Social Security Benefits
  + QAGEDEP = Percent Population under 5 years or 65 and over
  + MEDAGE = Median age 
* Housing Status 
  + PPUNIT = People per Unit (Average household size)
  + QFAM = Percent Children Living with both parents 
* Social Status
  + QCVLUN = Percent Unemployment for Civilian in Labor Force 16 Years and Over
  + QBLACK = Percent Black or African American Alone
  + QNOAUTO = Percent Housing Units with No Car
  + QPOVTY = Percent Poverty
* Gender
  + QFEMALE = Percent Female 
  + QFEMLBR = Percent Female Participation in Labor Force
  
  
# Variable Search

```{r v_search}

v19 <- load_variables(2019, "acs5", cache = TRUE)

```



```{r tidycensus}

#acs variables 
vars <- c(	
"QRICH" = "B19001_017", #need to divide households 
"households" = "B09019_002",
"MDHSEVAL" = 	"B25077_001",
"PERCAP" = "B19301_001",
"QESL-Spanish" = "B06007_005", #need to be added together and divided by pop 
"QESL-Other" = "B06007_008",
"QSPANISH" = "B03001_003", # need to divide by pop
"POP" = "B03001_001",
"QED12LES" = "B16010_002", # divide by pop
"QSSBEN" = "B19055_002", #divide by pop 
"QAGEDEP-under5" = "B06001_002",#add together and divide by pop
"QAGEDEP-over65" = "B18135_024",
"MEDAGE"= "B07002_001",
"PPUNIT" = "B25010_001",
"QFAM_under6" = "B05009_003", #add together and divide by children
"QFAM_to17" = "B05009_021",
"children" = "B05009_001",
"QCVLUN" = "B23025_005", # divide by pop 
"QBLACK" = "B18101B_001", #divide by pop,
"QNOAUTO" = "B08203_002", #divide by households 
"QPOVTY" = "B17020_002", #divide by pop
"QFEMALE" = "B01001_026", # divide by pop
"wlab1" = "C23002A_004", #can't really pull female labor participation except by race
"wlab2" = "C23002B_004", #Then divide by over 16
"wlab3" = "C23002C_004",
"wlab4" = "C23002D_004",
"wlab5" = "C23002E_004",
"wlab6" = "C23002F_004",
"wlab7" = "C23002G_004",
"f1" = "B01001_030", #pulling pop by age is actually the worst!
"f2" = "B01001_031",
"f3" = "B01001_032",
"f4" = "B01001_033",
"f5" = "B01001_034",
"f6" = "B01001_035",
"f7" = "B01001_036",
"f8" = "B01001_037",
"f9" = "B01001_038",
"f10" = "B01001_039",
"f11" = "B01001_040",
"f12" = "B01001_041",
"f13" = "B01001_042",
"f14" = "B01001_043",
"female_over65" = "B15001_076" 
)

counties <- c("Travis", "Bastrop","Blanco", "Burnet", "Caldwell",
              "Fayette", "Hays", "Lee", "Llano", "Williamson")

#enter your key census_api_key("Your key here")
acs <-get_acs(state="TX", county = counties, geography="tract",year = 2019,
              variables=vars, geometry= F) 

df_raw <- acs |> 
  select(GEOID, variable, estimate) |> 
  pivot_wider(id_cols = GEOID, 
              names_from = variable,
              values_from = estimate) 

```

# Data cleaning

```{r dc}

df <- df_raw |>
  mutate(
    QRICH = QRICH / households,
    QESL = (`QESL-Spanish` + `QESL-Other`) / POP,
    QSPANISH = QSPANISH / POP,
    QED12LES = QED12LES / POP,
    QSSBEN = QSSBEN / POP,
    QAGEDEP = (`QAGEDEP-under5` + `QAGEDEP-over65`) / POP,
    QFAM = (QFAM_under6 + QFAM_to17) / children,
    QCVLUN = QCVLUN / POP,
    QBLACK = QBLACK / POP,
    QNOAUTO = QNOAUTO / households,
    QPOVTY = QPOVTY / POP,
    QFEMALE = QFEMALE / POP,
    QFEMLBR = (wlab1 + wlab2 + wlab3 + wlab4 + wlab5 + wlab6 + wlab7) /
      (
        f1 + f2 + f3 + f4 + f5 + f6 + f7 + f8 + f9 + f10 + f11 + f12 + f13 + f14 + female_over65
      )
  ) |>
  select(
    -c(
      #Variables to remove
      POP,
      children,
      QFAM_under6,
      QFAM_to17,
      households,
      female_over65,
      `QAGEDEP-under5`,
      `QAGEDEP-over65`,
      `QESL-Spanish`,
      `QESL-Other`,
      starts_with("wlab"),
      "f1",
      "f2",
      "f3" ,
      "f4",
      "f5" ,
      "f6" ,
      "f7" ,
      "f8" ,
      "f9" ,
      "f10" ,
      "f11" ,
      "f12",
      "f13" ,
      "f14" ,
    )
  )

```


# Principal Components 

```{r pca}

pc <- prcomp(df[,-1],#have to take our geoid column
             center = TRUE,
            scale. = TRUE)

```

